---
title: "Computing overlap fractions between soundings and a fine grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing overlap fractions between soundings and a fine grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 5,
  fig.height = 5
)
```

`compute_overlap_fractions()` computes the change-of-support matrix $A$ where
entry $A_{ij}$ is the fraction of sounding $i$'s area covered by fine-grid
cell $j$:

$$A_{ij} = \frac{|D_i \cap D_j^{(\ell)}|}{|D_i|}$$

The denominator is always the **sounding area** $|D_i|$, not the cell area.
Row $i$ sums to 1 when the sounding lies entirely within the grid, and to less
than 1 when it extends beyond the grid boundary.

This vignette builds up from the simplest possible case to a more realistic
one, so the results can be verified by eye at each step.

## Setup

```{r libraries}
library(sf)
library(ggplot2)
library(spatintegrate)
```

```{r helpers}
# Build a single rectangular cell as an sfg
make_cell <- function(x0, y0, w = 1, h = w) {
  st_polygon(list(matrix(
    c(x0,   y0,
      x0+w, y0,
      x0+w, y0+h,
      x0,   y0+h,
      x0,   y0),
    ncol = 2, byrow = TRUE
  )))
}

# Build a regular grid of nrow x ncol unit cells
make_grid <- function(nrow, ncol, crs = 32619) {
  cells <- vector("list", nrow * ncol)
  k <- 1L
  for (r in 0:(nrow - 1)) {
    for (c in 0:(ncol - 1)) {
      cells[[k]] <- make_cell(c, r)
      k <- k + 1L
    }
  }
  st_sf(
    cell     = seq_len(nrow * ncol),
    geometry = st_sfc(cells, crs = crs)
  )
}

# Standard plot used across all examples
plot_overlap <- function(fine_grid, soundings, title, subtitle = NULL) {
  # Format labels: integers as "1", fractions as "0.417", zeros as "0"
  fmt <- function(x) {
    ifelse(x == 0, "0",
      ifelse(abs(x - round(x)) < 1e-9,
        as.character(as.integer(round(x))),
        as.character(round(x, 3))
      )
    )
  }

  ggplot() +
    geom_sf(
      data      = fine_grid,
      aes(fill  = fraction),
      colour    = "grey40",
      linewidth = 0.4
    ) +
    geom_sf(
      data      = soundings,
      fill      = NA,
      colour    = "#0F6E56",
      linewidth = 0.9,
      linetype  = "dashed"
    ) +
    geom_sf_text(
      data        = fine_grid,
      aes(label   = fmt(fraction)),
      size        = 3.5,
      colour      = "grey15"
    ) +
    scale_fill_distiller(
      palette   = "Greens",
      direction = 1,
      limits    = c(0, 1),
      name      = "Overlap\nfraction"
    ) +
    labs(
      title    = title,
      subtitle = if (!is.null(subtitle)) subtitle else
        "Each label: A[i,j] = |sounding \u2229 cell j| / |sounding|",
      x = NULL,
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid      = element_blank(),
      axis.text       = element_blank(),
      plot.title      = element_text(size = 11, face = "bold"),
      plot.subtitle   = element_text(size = 9, colour = "grey40"),
      legend.position = "right"
    )
}
```

---

## Example 1: sounding exactly partitioned into 4 equal cells

The simplest possible check. The sounding coincides exactly with the outer
boundary of a $2 \times 2$ grid of unit cells. Each cell takes up exactly
one quarter of the sounding, so every entry should be $0.25$ and the row
should sum to $1$.

```{r ex1-build}
fine_grid_1 <- make_grid(nrow = 2, ncol = 2)

sounding_1 <- st_sf(
  sounding = "full",
  geometry = st_sfc(make_cell(0, 0, w = 2, h = 2), crs = 32619)
)
```

```{r ex1-compute}
A1 <- compute_overlap_fractions(sounding_1, fine_grid_1)
A1
Matrix::rowSums(A1)
```

All four entries are $0.25$ and the row sums to $1$. ✓

```{r ex1-plot, fig.alt = "2x2 grid fully covered by sounding, all fractions 0.25"}
fine_grid_1$fraction <- as.numeric(A1[1, ])
plot_overlap(
  fine_grid_1, sounding_1,
  title    = "Example 1: sounding = full 2 \u00d7 2 grid",
  subtitle = "All four cells get exactly 1/4 of the sounding area"
)
```

---

## Example 2: sounding offset to one side (row sum = 1)

Now the sounding sits entirely within the grid but is not centred. We use
a $2.4 \times 1$ sounding over a $3 \times 1$ grid of unit cells, flush to
the left edge. Sounding area $= 2.4$.

The expected fractions are easy to work out:

| Cell  | Intersection | Area | Fraction ($\div\, 2.4$) |
|-------|--------------|------|-------------------------|
| left  | full cell    | 1.0  | $1/2.4 \approx$ **0.417** |
| mid   | full cell    | 1.0  | $1/2.4 \approx$ **0.417** |
| right | left 40%     | 0.4  | $0.4/2.4 \approx$ **0.167** |

Row sum: $0.417 + 0.417 + 0.167 = 1.0$ ✓

```{r ex2-build}
fine_grid_2 <- make_grid(nrow = 1, ncol = 3)

sounding_2 <- st_sf(
  sounding = "offset",
  geometry = st_sfc(make_cell(0, 0, w = 2.4, h = 1), crs = 32619)
)
```

```{r ex2-compute}
A2 <- compute_overlap_fractions(sounding_2, fine_grid_2)
round(A2, 3)
round(Matrix::rowSums(A2), 10)
```

```{r ex2-plot, fig.width = 6, fig.height = 2.5,
    fig.alt = "3x1 grid with sounding covering left 2.4 cells"}
fine_grid_2$fraction <- as.numeric(A2[1, ])
plot_overlap(
  fine_grid_2, sounding_2,
  title    = "Example 2: sounding partially overlaps right cell",
  subtitle = "Sounding area = 2.4; right cell is 40% covered so gets 40% of a full-cell share"
)
```

The dashed sounding outline visibly cuts $40\%$ of the way into the right
cell. Its fraction ($\approx 0.167$) is exactly $40\%$ of a full-cell share
($\approx 0.417$). ✓

---

## Example 3: sounding extends beyond the grid (row sum < 1)

When part of the sounding falls outside the grid the row sum drops below 1 —
the "missing" fraction is the portion of the sounding area with no latent
cells beneath it. This is the typical situation at the edges of an OCO-2
target-mode domain.

We use a $4 \times 4$ sounding centred on a $3 \times 3$ fine grid, so the
sounding overhangs by $0.5$ units on each side. Sounding area $= 16$; total
grid area $= 9$. Since the sounding is centred and rectangular, every grid
cell is fully enclosed, so all 9 entries equal $1/16 = 0.0625$ and the row
sums to $9/16 = 0.5625$.

```{r ex3-build}
fine_grid_3 <- make_grid(nrow = 3, ncol = 3)

# Centre the 4x4 sounding on the 3x3 grid
sounding_3 <- st_sf(
  sounding = "overhanging",
  geometry = st_sfc(make_cell(-0.5, -0.5, w = 4, h = 4), crs = 32619)
)
```

```{r ex3-compute}
A3 <- compute_overlap_fractions(sounding_3, fine_grid_3)
round(A3, 4)
round(Matrix::rowSums(A3), 6)  # 9/16 = 0.5625
```

```{r ex3-plot, fig.alt = "3x3 grid fully inside an overhanging sounding"}
fine_grid_3$fraction <- as.numeric(A3[1, ])
plot_overlap(
  fine_grid_3, sounding_3,
  title    = "Example 3: sounding extends beyond the grid",
  subtitle = paste0(
    "Row sum = ", round(Matrix::rowSums(A3), 4),
    " \u2248 9/16 \u2014 only 9/16 of the sounding area overlaps the grid"
  )
)
```

The sounding outline (dashed) extends well past all four edges. Every cell is
fully enclosed and equally covered, so all fractions are $1/16 = 0.0625$. The
row sum of $0.5625$ tells us that just over half the sounding's signal is
"seen" by the latent grid — the rest falls outside and is unmodelled. ✓

---

## Connecting to the downscaling model

`A` plugs directly into the Bayesian downscaling model as the linear map
from the latent fine-grid vector $r^{(\ell)}$ to the coarse observations:

$$y^{(o)} = X^{(o)} \beta + A\, r^{(\ell)} + \varepsilon$$

The sparsity of `A` is essential at scale. Each sounding overlaps only a
small number of fine-grid cells, so the vast majority of entries are zero.
At OCO-2 target-mode scale (6850 soundings, ~50 000 cells), sparsity exceeds
99%:

```{r sparsity, eval = FALSE}
cat("Dimensions:", nrow(A_large), "soundings x", ncol(A_large), "cells\n")
cat("Non-zero  :", Matrix::nnzero(A_large), "\n")
cat("Sparsity  :", round(1 - Matrix::nnzero(A_large) / prod(dim(A_large)), 3), "\n")
```

## Parallel execution

For large inputs, parallelise over soundings using any `future` backend:

```{r parallel, eval = FALSE}
library(future)
plan(multisession, workers = 4)

A_large <- compute_overlap_fractions(
  soundings = oco2_soundings,
  fine_grid  = boston_grid,
  parallel  = TRUE
)

plan(sequential)
```
