---
title: "integrate_basis: Numerical Integration over Polygons"
description: >
  How to use integrate_basis() to compute area-weighted averages of functions
  over sf polygons, with and without stratum boundaries.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{integrate_basis: Numerical Integration over Polygons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
library(spatintegrate)
library(sf)
library(ggplot2)
library(dplyr)
library(gridExtra)

theme_article <- function() {
  theme_minimal(base_size = 11) +
    theme(panel.border = element_rect(fill = NA, colour = "grey60"),
          legend.position = "none")
}
```

`integrate_basis()` is the main function of `spatintegrate`. Given a function
`basis_fn` and a set of polygons, it returns an $n \times k$ matrix $A$ where:

$$A_{ij} = \frac{1}{|D_i|} \int_{D_i} \phi_j(s) \, ds$$

Each row is one polygon, each column is one output of `basis_fn`. The integral
is approximated by triangulating each polygon and applying QMC quadrature with
`n_per_triangle` Sobol points per triangle.

---

## The `basis_fn` contract

`basis_fn` must accept an $n \times 2$ matrix of projected $(x, y)$ coordinates
and return an $n \times k$ matrix. The value of $k$ is inferred automatically
— you don't pass it.

```{r basis-contract}
# Gaussian bump centred at (cx, cy) with bandwidth sigma — k = 1
gaussian_basis <- function(coords, cx = 150, cy = 100, sigma = 80) {
  d2 <- (coords[, 1] - cx)^2 + (coords[, 2] - cy)^2
  matrix(exp(-d2 / (2 * sigma^2)), ncol = 1)
}

# Verify the contract: 3 points in, 3x1 matrix out
test_pts <- matrix(c(100,100,  200,50,  300,150), ncol = 2, byrow = TRUE)
gaussian_basis(test_pts)
```

This is the basis used throughout this vignette. Its integral over an arbitrary
polygon has no closed form — the QMC approximation has to do genuine work,
unlike a constant function which would be trivially exact regardless of
sampling density.

---

## Basic usage

```{r setup-polygons, fig.height = 3.5}
# Four overlapping parallelogram polygons in UTM 19N (meters)
make_parallelogram <- function(x0, y0, w, h, shear) {
  sf::st_polygon(list(matrix(c(
    x0,             y0,
    x0 + w,         y0,
    x0 + w + shear, y0 + h,
    x0 + shear,     y0 + h,
    x0,             y0
  ), ncol = 2, byrow = TRUE)))
}

polygons_sf <- sf::st_sf(
  id       = 1:4,
  geometry = sf::st_sfc(
    make_parallelogram(0,   0,   200, 100, 30),
    make_parallelogram(80,  20,  200, 100, 30),
    make_parallelogram(160, 40,  200, 100, 30),
    make_parallelogram(40,  80,  200, 100, 30)
  ),
  crs = 32619
)

poly_df <- do.call(bind_rows, lapply(1:4, function(i) {
  coords <- sf::st_coordinates(polygons_sf[i, ])
  data.frame(x = coords[,"X"], y = coords[,"Y"], id = factor(i))
}))

ggplot() +
  geom_polygon(data = poly_df, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.2, linewidth = 0.8) +
  scale_fill_manual(values = c("#2166ac","#d6604d","#4dac26","#8073ac")) +
  scale_colour_manual(values = c("#2166ac","#d6604d","#4dac26","#8073ac")) +
  coord_equal() +
  labs(title = "Four overlapping polygons", x = NULL, y = NULL) +
  theme_article()
```

The `gaussian_basis` defined above is used throughout. Values differ across
polygons because they sit at different distances from the bump centre.

```{r basic-integrate}
A <- integrate_basis(gaussian_basis, polygons_sf)
cat("Gaussian basis integrals (one per polygon):\n")
print(round(A, 6))
```

The values differ across polygons because they sit in different positions
relative to the bump centre — polygons closer to $(150, 100)$ accumulate
higher average values.

---

## Verifying integration accuracy

Without a closed-form answer, we use high-n Monte Carlo as a reference.
Uniform random sampling at large $n$ is an unbiased estimator, so its value
at $n = 100{,}000$ is a reliable benchmark:

```{r accuracy, fig.height = 3.5}
# Monte Carlo reference at high n (unbiased, converges slowly)
mc_reference <- function(polygon_sfc, n = 100000) {
  bbox   <- sf::st_bbox(polygon_sfc)
  pts    <- matrix(c(
    runif(n, bbox["xmin"], bbox["xmax"]),
    runif(n, bbox["ymin"], bbox["ymax"])
  ), ncol = 2)
  pts_sfc <- sf::st_sfc(lapply(seq_len(n), function(i) sf::st_point(pts[i,])),
                         crs = sf::st_crs(polygon_sfc))
  inside  <- as.vector(sf::st_within(pts_sfc, polygon_sfc, sparse = FALSE))
  if (sum(inside) == 0) return(NA_real_)
  mean(gaussian_basis(pts[inside, , drop = FALSE]))
}

set.seed(42)
reference <- sapply(seq_len(nrow(polygons_sf)), function(i) {
  mc_reference(sf::st_geometry(polygons_sf)[i])
})

A_qmc <- integrate_basis(gaussian_basis, polygons_sf, n_per_triangle = 32)

results <- data.frame(
  polygon   = 1:4,
  QMC_n16   = round(integrate_basis(gaussian_basis, polygons_sf, n_per_triangle = 16)[,1], 6),
  QMC_n32   = round(A_qmc[, 1], 6),
  MC_ref    = round(reference, 6)
)
knitr::kable(results, caption = "QMC vs Monte Carlo reference (n = 100,000)")
```

QMC at `n_per_triangle = 32` matches the high-n Monte Carlo reference closely,
and it got there with far fewer function evaluations.

---

## `n_per_triangle`: accuracy vs cost

```{r n_per_triangle, fig.height = 3.5}
ns     <- c(4, 8, 16, 32, 64, 128)
errors <- sapply(ns, function(n) {
  A <- integrate_basis(gaussian_basis, polygons_sf, n_per_triangle = n)
  mean(abs(A[, 1] - reference))
})

err_df <- data.frame(n_per_triangle = ns, mean_abs_error = errors)

ggplot(err_df, aes(n_per_triangle, mean_abs_error)) +
  geom_line(colour = "#2166ac", linewidth = 0.8) +
  geom_point(colour = "#2166ac", size = 2.5) +
  scale_x_log10(breaks = ns) +
  scale_y_log10() +
  labs(title    = "Integration error vs n_per_triangle",
       subtitle = "Mean absolute error vs Monte Carlo reference (n = 100,000)",
       x = "n_per_triangle", y = "Mean absolute error (log scale)") +
  theme_article() +
  theme(legend.position = "none")
```

Error falls roughly as $O(n^{-1})$ for QMC — faster than the $O(n^{-1/2})$
of plain Monte Carlo. For most smooth basis functions `n_per_triangle = 16`
(the default) is more than sufficient.

---

## Strata

When `basis_fn` has discontinuities at known boundaries — e.g. a covariance
kernel that treats two regions as uncorrelated — supply `strata_sf`. Each
polygon is intersected with each stratum before triangulating, guaranteeing
that no QMC point straddles a boundary.

Here the stratum is a small circle sitting inside the scene. Stratum A is the
circle, stratum B is everything else.

```{r strata-setup, fig.height = 4}
# Small circle stratum centred at (150, 80), radius 60m
circle <- sf::st_buffer(
  sf::st_sfc(sf::st_point(c(150, 80)), crs = 32619),
  dist = 60
)
domain   <- sf::st_as_sfc(sf::st_bbox(
  c(xmin = -50, ymin = -50, xmax = 450, ymax = 250),
  crs = 32619
))
strata_sf <- sf::st_sf(
  name     = c("circle", "outside"),
  geometry = sf::st_sfc(
    sf::st_geometry(circle)[[1]],
    sf::st_geometry(sf::st_difference(domain, circle))[[1]],
    crs = 32619
  )
)

# Visualise polygons + stratum
circle_df <- as.data.frame(sf::st_coordinates(circle))

ggplot() +
  geom_polygon(data = poly_df, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.15, linewidth = 0.8) +
  geom_polygon(data = circle_df, aes(X, Y),
               fill = "#d6604d", alpha = 0.35,
               colour = "#d6604d", linewidth = 1) +
  scale_fill_manual(values = c("#2166ac","#d6604d","#4dac26","#8073ac")) +
  scale_colour_manual(values = c("#2166ac","#d6604d","#4dac26","#8073ac")) +
  coord_equal() +
  labs(title    = "Polygons with circular stratum (red)",
       subtitle = "QMC points will be confined within or outside the circle",
       x = NULL, y = NULL) +
  theme_article()
```

```{r strata-integrate}
A_strat <- integrate_basis(gaussian_basis, polygons_sf,
                            strata_sf      = strata_sf,
                            n_per_triangle = 32)

cat("With strata — Gaussian integrals:\n")
print(round(A_strat, 6))

cat("\nWithout strata:\n")
print(round(A_qmc, 6))

cat("\nDifference (should be small — same integral, different sampling):\n")
print(round(A_strat - A_qmc, 6))
```

The results with and without strata should be nearly identical — both are
approximating the same integral. The difference is that with strata, no QMC
point lies on the circle boundary, which matters when `basis_fn` treats the
inside and outside as categorically different regions.

```{r strata-plot, fig.height = 4}
# Show the intersection pieces for polygon 1
pieces <- lapply(seq_len(nrow(strata_sf)), function(i) {
  inter <- suppressWarnings(
    sf::st_intersection(sf::st_geometry(polygons_sf)[1],
                        sf::st_geometry(strata_sf)[i])
  )
  if (length(inter) == 0 || sf::st_is_empty(inter)) return(NULL)
  data.frame(sf::st_coordinates(inter), stratum = strata_sf$name[i])
})
pieces_df <- do.call(rbind, Filter(Negate(is.null), pieces))

ggplot() +
  geom_polygon(data = poly_df |> filter(id == 1), aes(x, y),
               fill = NA, colour = "black", linewidth = 1) +
  geom_polygon(data = pieces_df, aes(X, Y, fill = stratum, group = L2),
               alpha = 0.4, colour = "grey30", linewidth = 0.5) +
  scale_fill_manual(values = c(circle = "#d6604d", outside = "#a6cee3")) +
  coord_equal() +
  labs(title = "Polygon 1 intersected with strata",
       fill = "Stratum", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.border = element_rect(fill = NA, colour = "grey60"))
```

---

## What `basis_fn` receives

It's worth being explicit about the coordinate system. `basis_fn` receives
coordinates in the **same projected CRS as `polygons_sf`** — in meters for a
UTM projection. Your basis function must be defined in that space.

```{r coords-demo}
# Capture what basis_fn actually receives for polygon 1
captured_coords <- NULL
capture_basis <- function(coords) {
  captured_coords <<- coords
  matrix(1, nrow = nrow(coords), ncol = 1)
}
integrate_basis(capture_basis, polygons_sf[1, ], n_per_triangle = 4)

cat("Coordinates received by basis_fn (first 4 rows):\n")
print(head(captured_coords, 4))
cat("CRS: UTM 19N — units are meters\n")
```

---

## Parallelism

For large numbers of polygons, set `parallel_plan = "ambient"` and configure
a `future` plan beforehand:

```{r parallel, eval = FALSE}
library(future)
plan(multisession, workers = 4)

A <- integrate_basis(gaussian_basis, polygons_sf, parallel_plan = "ambient")

plan(sequential)  # reset when done
```

The QMC sequence is generated once and shared across all workers, so results
are identical to sequential execution.

---

## Summary

| Argument | Default | Purpose |
|---|---|---|
| `basis_fn` | required | Function: n×2 coords → n×k matrix |
| `polygons_sf` | required | sf object of polygons to integrate over |
| `strata_sf` | `NULL` | Optional strata for discontinuous basis functions |
| `n_per_triangle` | `16` | QMC points per triangle — controls accuracy |
| `min_area` | `0` | Drop triangles at or below this area |
| `parallel_plan` | `"sequential"` | `"ambient"` to use `future::plan()` |

The key constraint: `polygons_sf` (and `strata_sf` if supplied) must be in a
**projected CRS**. Use `ensure_projected()` to convert geographic inputs first.
