---
title: "Comparing SAR-Raster and Fourier-SAR Downscaling"
description: >
  A practitioner-oriented comparison of SAR-raster and Fourier-SAR
  representations for Bayesian downscaling, showing how the same SAR
  parameters can be translated into a spectral Fourier basis.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing SAR-Raster and Fourier-SAR Downscaling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)

library(spatintegrate)
library(fastblm)
library(sf)
library(Matrix)
library(spdep)
library(ggplot2)

theme_vignette <- function() {
  theme_minimal(base_size = 11) +
    theme(panel.border = element_rect(fill = NA, colour = "grey60"))
}

rmse <- function(x, truth) {
  sqrt(mean((x - truth)^2, na.rm = TRUE))
}

make_rect_domain <- function(domain_bbox, crs) {
  sf::st_as_sfc(
    sf::st_bbox(
      setNames(as.numeric(domain_bbox), c("xmin", "ymin", "xmax", "ymax")),
      crs = crs
    )
  )
}

make_rook_W <- function(grid_sf) {
  # Neumann-reflection W: the unique W whose eigenvectors are the 2D
  # Neumann cosine functions cos((k1-1)*pi*i/m)*cos((k2-1)*pi*j/m),
  # with eigenvalues (cos(pi*(k1-1)/m) + cos(pi*(k2-1)/m))/2.
  #
  # Construction: each cell has weight 1/4 per neighbor (off-diagonal),
  # plus 1/4 added to the diagonal for each MISSING neighbor (Neumann
  # reflection: the ghost cell outside the boundary equals the boundary
  # cell itself). Equivalently:
  #   diagonal entry = (4 - number of real neighbors) / 4
  #   off-diagonal   = 1/4 for each real neighbor
  # This makes W symmetric and row-sums = 1 everywhere.
  nb <- spdep::poly2nb(grid_sf, queen = FALSE)
  n  <- length(nb)

  # Off-diagonal entries: weight 1/4 per real neighbor
  i_off <- rep(seq_len(n), lengths(nb))
  j_off <- unlist(nb, use.names = FALSE)
  if (length(i_off) == 0L) stop("No neighboring grid cells were found.")

  # Diagonal entries: (4 - n_neighbors) / 4  [Neumann reflection contribution]
  n_nb    <- lengths(nb)
  i_diag  <- seq_len(n)
  j_diag  <- seq_len(n)
  w_diag  <- (4L - n_nb) / 4

  # Combine
  i_all <- c(i_off, i_diag[w_diag > 0])
  j_all <- c(j_off, j_diag[w_diag > 0])
  w_all <- c(rep(1/4, length(i_off)), w_diag[w_diag > 0])

  Matrix::sparseMatrix(i = i_all, j = j_all, x = w_all,
                       dims = c(n, n))
}

make_sar_precision <- function(grid_sf, rho, tau2) {
  if (!is.numeric(rho) || length(rho) != 1L || rho <= 0 || rho >= 1)
    stop("`rho` must be a scalar in (0, 1).")
  if (!is.numeric(tau2) || length(tau2) != 1L || tau2 <= 0)
    stop("`tau2` must be a positive scalar.")
  W <- make_rook_W(grid_sf)
  I <- Matrix::Diagonal(nrow(W))
  Matrix::crossprod(I - rho * W) / tau2
}

# Evaluate the Neumann cosine basis at arbitrary point locations.
# Matches fourier_integrate_basis() exactly:
#   phi_j(s) = c_j * cos((j1-1)*pi*s_x/Lx) * cos((j2-1)*pi*s_y/Ly)
# coords   : n x 2 matrix of (x, y) in CRS units
# freq_grid: list from fourier_freq_grid()
fourier_evaluate_basis <- function(coords, freq_grid) {
  bbox  <- freq_grid$domain_bbox
  Lx    <- bbox[3] - bbox[1]
  Ly    <- bbox[4] - bbox[2]
  ox    <- freq_grid$omega_mat[, 1]   # (j1-1)*pi/Lx  for each basis fn
  oy    <- freq_grid$omega_mat[, 2]   # (j2-1)*pi/Ly
  nc    <- freq_grid$norm_const

  sx <- as.matrix(coords)[, 1]
  sy <- as.matrix(coords)[, 2]

  # outer products: each row is one location, each col is one basis fn
  cos_x <- cos(outer(sx, ox))   # n x r
  cos_y <- cos(outer(sy, oy))   # n x r

  sweep(cos_x * cos_y, 2, nc, `*`)   # apply norm_const column-wise
}
```

This vignette compares two ways of fitting the same downscaling model:

1. a **SAR-raster model**, where the latent field is represented by one value per
   grid cell; and

2. a **Fourier-SAR model**, where the latent field is represented by Fourier
   basis coefficients, but the prior is parameterized using familiar SAR
   parameters.

The main message is that the observation model is unchanged. In both cases,

\[
y = A\theta + \epsilon.
\]

What changes is the meaning of \(\theta\) and the construction of \(A\).

In the SAR-raster model, \(\theta\) is a vector of grid-cell values and \(A\)
contains area-overlap fractions. In the Fourier-SAR model, \(\theta\) is a vector
of Fourier coefficients and \(A\) contains exact footprint averages of Neumann
cosine basis functions.

The prior is also translated. The Fourier-SAR prior takes SAR parameters
\((\rho, \tau^2)\), converts them internally to the matched Matérn/SPDE
parameterization, and constructs the corresponding diagonal spectral precision.

---

## SAR and Fourier as two coordinate systems

A SAR model encodes smoothness locally, through neighboring grid cells. A common
rook-neighbor SAR precision has the form

\[
Q_{\mathrm{SAR}}
=
(I - \rho W)^\top (I - \rho W) / \tau^2,
\]

where \(W\) is a neighborhood matrix, \(\rho\) controls spatial dependence, and
\(\tau^2\) controls prior scale.

A Fourier-SAR model encodes the same idea spectrally. Low-frequency basis
functions represent broad, smooth spatial variation. High-frequency basis
functions represent rapidly oscillating spatial variation. The SAR-matched
spectral prior gives larger prior variance to smooth low-frequency components
and smaller prior variance to rough high-frequency components.

A useful translation is:

| SAR-raster view | Fourier-SAR view |
|---|---|
| Grid-cell values | Fourier coefficients |
| Neighbor smoothing | Frequency-dependent shrinkage |
| Sparse precision matrix | Diagonal spectral precision |
| Refining the raster | Adding frequencies (increasing J) |
| \(A\) averages pixels | \(A\) averages basis functions |
| Prediction tied to fitted raster | Prediction at arbitrary points or regions |

Thus, the Fourier formulation should be read as a change of basis, not as a
different downscaling model.

---

## Step 1: create a synthetic latent field

```{r make-domain}
# Domain sized to observations: 5km obs window + 2.5km buffer each side = 10km
domain_bbox <- c(0, 0, 10e3, 10e3)
names(domain_bbox) <- c("xmin", "ymin", "xmax", "ymax")

grid_n <- 120
x_seq <- seq(domain_bbox["xmin"], domain_bbox["xmax"], length.out = grid_n)
y_seq <- seq(domain_bbox["ymin"], domain_bbox["ymax"], length.out = grid_n)

truth_df <- expand.grid(x = x_seq, y = y_seq)

# Field varies on scale of the domain
truth_df$truth <-
  1.0 * sin(2 * pi * truth_df$x / 10e3) +
  0.7 * cos(2 * pi * truth_df$y / 10e3) +
  0.4 * sin(2 * pi * (truth_df$x + truth_df$y) / 6e3)

truth_sf <- sf::st_as_sf(truth_df, coords = c("x", "y"), crs = 32619)

ggplot(truth_df, aes(x, y, fill = truth)) +
  geom_raster() +
  coord_equal() +
  labs(title = "Synthetic latent field",
       x = "Easting (m)", y = "Northing (m)", fill = "v(s)") +
  theme_vignette()
```

---

## Step 2: create synthetic satellite footprints

```{r make-footprints}
make_footprint <- function(cx, cy, width = 1300, height = 2300, angle_deg = 15) {
  angle <- angle_deg * pi / 180
  hw <- width / 2; hh <- height / 2
  corners <- matrix(c(-hw,-hh, hw,-hh, hw,hh, -hw,hh, -hw,-hh),
                    ncol = 2, byrow = TRUE)
  rot <- matrix(c(cos(angle), -sin(angle), sin(angle), cos(angle)), nrow = 2)
  rotated <- corners %*% t(rot)
  rotated[, 1] <- rotated[, 1] + cx
  rotated[, 2] <- rotated[, 2] + cy
  sf::st_polygon(list(rotated))
}

set.seed(1)
n_soundings <- 250
# 5km x 5km cluster centred in domain (2.5km - 7.5km), 2.5km buffer to boundary
track_x <- runif(n_soundings, 2.5e3, 7.5e3)
track_y <- runif(n_soundings, 2.5e3, 7.5e3)

soundings_sf <- sf::st_sf(
  id = seq_len(n_soundings),
  geometry = sf::st_sfc(mapply(make_footprint, track_x, track_y, SIMPLIFY = FALSE),
                         crs = 32619)
)

ggplot() +
  geom_sf(data = soundings_sf, fill = "skyblue", alpha = 0.35, colour = "grey30") +
  coord_sf() +
  labs(title = "Synthetic satellite footprints",
       subtitle = "Centroids sampled densely in 2D, producing overlap in both directions",
       x = "Easting (m)", y = "Northing (m)") +
  theme_vignette()
```

---

## Step 3: simulate footprint-level observations

```{r simulate-observations}
average_truth_over_footprint <- function(i) {
  inside <- sf::st_within(truth_sf, soundings_sf[i, ], sparse = FALSE)[, 1]
  if (!any(inside)) return(NA_real_)
  mean(truth_df$truth[inside])
}

mu   <- vapply(seq_len(n_soundings), average_truth_over_footprint, numeric(1))
keep <- !is.na(mu)
soundings_sf <- soundings_sf[keep, ]
mu   <- mu[keep]
n_soundings  <- length(mu)

sigma_eps <- 0.15
y <- mu + rnorm(n_soundings, sd = sigma_eps)

soundings_sf$y  <- y
soundings_sf$mu <- mu

cat("Usable soundings:", n_soundings, "\n")
cat("Noise sd:", sigma_eps, "\n")
```

---

## Step 4: build the SAR-raster design matrix

```{r sar-grid}
domain_poly <- make_rect_domain(domain_bbox, crs = sf::st_crs(soundings_sf))

sar_grid_sf <- sf::st_sf(
  cell_id  = seq_along(sf::st_make_grid(domain_poly, n = c(80, 80), square = TRUE)),
  geometry = sf::st_make_grid(domain_poly, n = c(80, 80), square = TRUE)
)

n_cells <- nrow(sar_grid_sf)
delta   <- sqrt(as.numeric(sf::st_area(sar_grid_sf[1, ])))
cat("SAR raster cells:", n_cells, "(", sqrt(n_cells), "x", sqrt(n_cells), ")\n")
cat("Cell size (delta):", delta, "m\n")
cat("Footprint/cell ratio: ~", round(1300/delta, 1), "x", round(2300/delta, 1), "\n")
```

```{r sar-A}
A_sar <- compute_overlap_fractions(
  soundings = soundings_sf,
  fine_grid = sar_grid_sf,
  sparse = TRUE
)

cat("Dimensions of SAR A:", nrow(A_sar), "x", ncol(A_sar), "\n")
cat("Range of SAR row sums:", range(Matrix::rowSums(A_sar)), "\n")
```

---

## Step 5: build the SAR prior

```{r sar-prior}
rho  <- 0.95
tau2 <- 1.0    # SAR conditional variance -- same parameter for both models
phi  <- 1.0

cat("SAR parameters: delta =", round(delta,2), " rho =", rho, " tau2 =", tau2, "\n")

Q_sar <- make_sar_precision(grid_sf = sar_grid_sf, rho = rho, tau2 = tau2)
cat("SAR precision:", nrow(Q_sar), "x", ncol(Q_sar),
    " nonzeros:", length(Q_sar@x), "\n")
```

```{r fit-sar}
fit_sar    <- fit_fastblm(y = y, A = A_sar, Q = Q_sar, phi = phi, solver = "cholesky")
theta_sar  <- fit_sar$posterior_mean
fitted_sar <- as.numeric(A_sar %*% theta_sar)
```

---

## Step 6: build the matched Fourier-SAR design matrix

The Fourier basis here is the **Neumann cosine basis** from the dissertation:

\[
\phi_{\mathbf{j}}(s)
= \prod_{k=1}^2 \sqrt{2 - \delta_{j_k,1}}\,
  \cos\!\big((j_k-1)\pi s_k / L_k\big),
\qquad j_k \in \{1,\ldots,J\}.
\]

`J` controls the resolution: `J = 1` gives only the constant; `J = 10` gives up
to 9 half-waves across the domain in each direction, for \(J^2 = 100\) basis
functions total.

```{r fourier-frequencies}
J <- 80

fg <- fourier_freq_grid(J = J, domain_bbox = domain_bbox)

cat("J =", J, "\n")
cat("Basis functions:", nrow(fg$indices), "= J^2\n")
cat("Columns in A:", nrow(fg$indices), "(one per basis function, cosine only)\n")
```

```{r fourier-A}
A_fourier <- fourier_integrate_basis(polygons_sf = soundings_sf, freq_grid = fg)

cat("Dimensions of Fourier A:", nrow(A_fourier), "x", ncol(A_fourier), "\n")
```

No columns are dropped — unlike a complex exponential basis, the Neumann cosine
basis has no redundant zero-sine column.

---

## Step 7: build the matched Fourier-SAR prior

```{r fourier-prior}
Q_fun_fourier <- fourier_sar_Q_fun(freq_grid = fg, delta = delta)

Q_fourier <- Q_fun_fourier(c(logit_rho = qlogis(rho), log_tau2 = log(tau2)))$Q

cat("Dimensions of Fourier Q:", nrow(Q_fourier), "x", ncol(Q_fourier), "\n")

summarize_prior(Q_fun = Q_fun_fourier,
                theta = c(logit_rho = qlogis(rho), log_tau2 = log(tau2)))
```

```{r fit-fourier}
fit_fourier    <- fit_fastblm(y = y, A = A_fourier, Q = Q_fourier,
                              phi = phi, solver = "cholesky")
theta_fourier  <- fit_fourier$posterior_mean
fitted_fourier <- as.numeric(A_fourier %*% theta_fourier)
```

---

## Step 8: compare fitted values at the observation scale

```{r compare-observation-scale}
obs_compare <- data.frame(y = y, truth = mu,
                          sar = fitted_sar, fourier = fitted_fourier)

knitr::kable(
  data.frame(
    model = c("SAR-raster", "Fourier-SAR"),
    rmse_against_truth = c(rmse(fitted_sar, mu), rmse(fitted_fourier, mu)),
    rmse_against_y     = c(rmse(fitted_sar, y),  rmse(fitted_fourier, y))
  ),
  digits = 4, caption = "Observation-scale comparison."
)
```

```{r plot-fitted-comparison}
ggplot(obs_compare, aes(sar, fourier)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  coord_equal() +
  labs(title = "Fitted footprint means: SAR-raster vs Fourier-SAR",
       x = "SAR-raster fitted value", y = "Fourier-SAR fitted value") +
  theme_vignette()
```

---

## Step 9: compare predictions on a common grid

For the SAR-raster model, the fitted coefficients are the grid-cell predictions
directly. For the Fourier-SAR model, we evaluate the basis at grid centroids.

```{r prediction-grid}
pred_pts <- sf::st_centroid(sar_grid_sf)
pred_xy  <- sf::st_coordinates(pred_pts)

pred_sar     <- as.numeric(theta_sar)
Phi_fourier  <- fourier_evaluate_basis(coords = pred_xy, freq_grid = fg)
pred_fourier <- as.numeric(Phi_fourier %*% theta_fourier)

nearest_idx <- FNN::get.knnx(
  data  = as.matrix(truth_df[, c("x","y")]),
  query = pred_xy, k = 1
)$nn.index[, 1]

pred_df <- data.frame(
  x          = pred_xy[, 1],
  y          = pred_xy[, 2],
  truth      = truth_df$truth[nearest_idx],
  sar        = pred_sar,
  fourier    = pred_fourier,
  difference = pred_fourier - pred_sar
)
```

```{r plot-predictions}
ggplot(pred_df, aes(x, y, fill = sar)) +
  geom_raster() + coord_equal() +
  labs(title = "SAR-raster posterior mean",
       x = "Easting (m)", y = "Northing (m)", fill = "mean") +
  theme_vignette()

ggplot(pred_df, aes(x, y, fill = fourier)) +
  geom_raster() + coord_equal() +
  labs(title = "Fourier-SAR posterior mean",
       x = "Easting (m)", y = "Northing (m)", fill = "mean") +
  theme_vignette()

ggplot(pred_df, aes(x, y, fill = difference)) +
  geom_raster() + coord_equal() +
  labs(title = "Difference: Fourier-SAR minus SAR-raster",
       x = "Easting (m)", y = "Northing (m)", fill = "diff") +
  theme_vignette()
```

```{r prediction-summary}
knitr::kable(
  data.frame(
    model = c("SAR-raster", "Fourier-SAR"),
    rmse_against_truth = c(rmse(pred_df$sar, pred_df$truth),
                           rmse(pred_df$fourier, pred_df$truth))
  ),
  digits = 4, caption = "Prediction-grid comparison."
)
```

---

## Step 10: sensitivity to Fourier truncation

```{r fourier-sensitivity}
J_values <- c(10, 20, 40, 60, 80)

sens <- lapply(J_values, function(J_now) {
  fg_now <- fourier_freq_grid(J = J_now, domain_bbox = domain_bbox)
  A_now  <- fourier_integrate_basis(polygons_sf = soundings_sf, freq_grid = fg_now)

  Q_fun_now <- fourier_sar_Q_fun(freq_grid = fg_now, delta = delta)
  Q_now     <- Q_fun_now(c(logit_rho = qlogis(rho), log_tau2 = log(tau2)))$Q

  fit_now      <- fit_fastblm(y = y, A = A_now, Q = Q_now,
                              phi = phi, solver = "cholesky")
  fitted_now   <- as.numeric(A_now %*% fit_now$posterior_mean)

  data.frame(J              = J_now,
             n_basis        = ncol(A_now),
             rmse_vs_truth  = rmse(fitted_now, mu),
             rmse_vs_sar    = rmse(fitted_now, fitted_sar))
})

sens_df <- do.call(rbind, sens)

knitr::kable(sens_df, digits = 4,
             caption = "Sensitivity to the number of basis functions (J^2 total).")
```

```{r plot-fourier-sensitivity}
ggplot(sens_df, aes(n_basis, rmse_vs_sar)) +
  geom_line() + geom_point() +
  labs(title = "Fourier-SAR approximation stabilizes as J increases",
       x = "Number of basis functions (J^2)",
       y = "RMSE against SAR-raster fitted values") +
  theme_vignette()
```

---

## Summary

Both models fit the same observation equation \(y = A\theta + \epsilon\). The
difference is representational: \(\theta\) is grid-cell values in the SAR-raster
model and Neumann cosine coefficients in the Fourier-SAR model.

The prior translation is direct via the exact discrete-Laplacian correspondence:
the same \((\rho, \tau^2, \delta)\) that defines the SAR-raster prior also defines
the Fourier-SAR spectral precision, and both can be read off in the other
parameterization via `summarize_prior()`.

The Fourier representation's main advantage is that the fitted coefficients can
be reused for prediction at new points, new grids, or new regions without
refitting.

```{r debug-dc}
cat("Mean of y:", mean(y), "\n")
cat("Mean of SAR fitted:", mean(fitted_sar), "\n")
cat("Mean of Fourier fitted:", mean(fitted_fourier), "\n")
cat("Offset (Fourier - SAR):", mean(fitted_fourier) - mean(fitted_sar), "\n\n")

# DC coefficient in each model
cat("SAR theta[1] (first cell, near DC):", theta_sar[1], "\n")
cat("Fourier theta for phi_1_1 (DC):", theta_fourier[colnames(A_fourier) == "phi_1_1"], "\n\n")

# Prior variance comparison at non-DC vs DC
Q_diag <- diag(Q_fourier)
cat("Fourier DC prior variance:", 1/Q_diag[colnames(A_fourier) == "phi_1_1"], "\n")
cat("Fourier non-DC prior variance range:", range(1/Q_diag[colnames(A_fourier) != "phi_1_1"]), "\n\n")

# What does the SAR Q give for the constant eigenvector?
ones <- rep(1, nrow(Q_sar))
sar_dc_precision <- as.numeric(t(ones) %*% Q_sar %*% ones) / nrow(Q_sar)^2
cat("SAR effective DC precision (per-cell):", sar_dc_precision, "\n")
cat("SAR effective DC prior variance:", 1/sar_dc_precision, "\n")
```
