---
title: "Exact Fourier Integration over Satellite Footprints"
description: >
  A step-by-step walkthrough of fourier_integrate_basis(): visualising
  soundings, computing the A matrix exactly, verifying the result, and
  reading off SAR-equivalent parameters.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exact Fourier Integration over Satellite Footprints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
library(spatintegrate)
library(sf)
library(ggplot2)

theme_vignette <- function() {
  theme_minimal(base_size = 11) +
    theme(panel.border = element_rect(fill = NA, colour = "grey60"))
}
```

This vignette walks slowly through what `fourier_integrate_basis()` does and
why. There is no model fitting here — the goal is just to understand the $A$
matrix.

---

## The setup

We want to estimate a spatial field $v(s)$ from satellite observations. Each
satellite sounding observes a **footprint** — a polygon on the ground — and
returns a single number that is approximately the area-averaged field value over
that footprint:

$$y_i \approx \frac{1}{|D_i|} \int_{D_i} v(s) \, ds$$

We represent the unknown field $v(s)$ as a linear combination of Fourier basis
functions:

$$v(s) \approx \sum_k \alpha_k \phi_k(s)$$

Plugging in, each observation becomes a linear function of the coefficients
$\alpha$:

$$y_i \approx \sum_k \alpha_k \underbrace{\frac{1}{|D_i|} \int_{D_i} \phi_k(s) \, ds}_{A_{ik}}$$

The matrix $A$ is what `fourier_integrate_basis()` computes. Each entry
$A_{ik}$ is the area-averaged value of basis function $k$ over footprint $i$.

---

## Step 1: make some soundings

We'll simulate a small set of OCO-2-like soundings over a 200km $\times$ 200km
domain. Each sounding is a small parallelogram (the satellite footprint is
roughly 1.3km $\times$ 2.3km, slightly tilted along the orbit track).

```{r make-soundings}
# Helper: one tilted footprint centred at (cx, cy)
make_footprint <- function(cx, cy, width = 1300, height = 2300, angle_deg = 15) {
  angle <- angle_deg * pi / 180
  # Four corners relative to centre, then rotate
  hw <- width  / 2
  hh <- height / 2
  corners <- matrix(c(-hw, -hh,
                       hw, -hh,
                       hw,  hh,
                      -hw,  hh,
                      -hw, -hh), ncol = 2, byrow = TRUE)
  rot <- matrix(c(cos(angle), -sin(angle),
                  sin(angle),  cos(angle)), nrow = 2)
  rotated <- corners %*% t(rot)
  rotated[, 1] <- rotated[, 1] + cx
  rotated[, 2] <- rotated[, 2] + cy
  sf::st_polygon(list(rotated))
}

# Simulate a small orbit track: 8 soundings spaced 1.8km apart (footprint
# height is 2.3km, so consecutive soundings overlap by ~500m)
set.seed(42)
n_soundings <- 8
cx <- 100e3 + seq(0, 0, length.out = n_soundings)
cy <- 100e3 + seq(0, (n_soundings - 1) * 1800, length.out = n_soundings)

soundings_sf <- sf::st_sf(
  id       = seq_len(n_soundings),
  geometry = sf::st_sfc(
    mapply(make_footprint, cx, cy, SIMPLIFY = FALSE),
    crs = 32619   # UTM 19N, units = meters
  )
)
```

Let's look at them.

```{r plot-soundings, fig.height = 5}
# Zoom to the sounding extent with a small buffer
buf <- 3000   # 3km buffer around the footprints
bb  <- sf::st_bbox(soundings_sf)

ggplot() +
  geom_sf(data = soundings_sf, aes(fill = factor(id)), alpha = 0.6,
          colour = "black", linewidth = 0.5) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  coord_sf(xlim = c(bb["xmin"] - buf, bb["xmax"] + buf),
           ylim = c(bb["ymin"] - buf, bb["ymax"] + buf)) +
  labs(title    = "Simulated satellite soundings",
       subtitle = sprintf("%d footprints, ~1.3km x 2.3km, tilted 15°, spaced 1.8km apart",
                          n_soundings),
       x = "Easting (m)", y = "Northing (m)") +
  theme_vignette()
```

Consecutive footprints overlap by about 500m — the footprint height is 2.3km
but they are spaced only 1.8km apart along the orbit track.

---

## Step 2: build the frequency grid

Think of the frequency grid as dividing the domain into $k$ equal strips in
each direction. If the domain is $L$ meters wide and you want $k$ complete
cycles of a sinusoid to fit across it, the corresponding spatial period is
$L/k$ — one full wave every $L/k$ meters. `K = 5` means we include up to 5
complete cycles across the domain in each direction, so the finest features we
can represent are $L/5$ wide.

The internal unit is angular frequency $\omega = 2\pi k / L$ (radians per
meter), which is just $2\pi$ times cycles-per-meter. The two are equivalent;
the angular form is more convenient for the integration formula.

`fourier_freq_grid()` takes:

- `K` — the maximum number of complete cycles across the domain in each
  direction. Total frequencies: $(2K+1)^2$ (positive, negative, and zero in
  each dimension).
- `domain_bbox` — the bounding box, used to set the period $L_x$ and $L_y$.

```{r freq-grid}
bbox  <- c(0, 0, 200e3, 200e3)   # xmin, ymin, xmax, ymax in meters
K     <- 5
freqs <- fourier_freq_grid(K = K, domain_bbox = bbox)

cat("Domain width: 200 km\n")
cat("K =", K, "=> finest resolvable feature: 200/5 = 40 km\n")
cat("Number of frequency pairs:", nrow(freqs), "= (2*5+1)^2\n")
cat("Columns in A:", 2 * nrow(freqs), "(cos and sin for each frequency)\n")
cat("\nFrequency grid in cycles per km (k / L):\n")
Lkm <- 200
print(head(round(freqs * Lkm / (2 * pi), 3)))   # convert to cycles/km
```

The `k` column values are just the integer indices $-K, \ldots, K$ — exactly
the same as dividing the domain into $2K$ equal intervals. Each row is one
$(k_x, k_y)$ pair, and the corresponding basis functions are
$\cos(2\pi k_x s_x / L_x) \cos(2\pi k_y s_y / L_y)$ and so on.

```{r plot-freqs, fig.height = 4}
freqs_df       <- as.data.frame(freqs * 200 / (2 * pi))   # convert to cycles/km
names(freqs_df) <- c("kx", "ky")
freqs_df$zero  <- freqs_df$kx == 0 & freqs_df$ky == 0

ggplot(freqs_df, aes(kx, ky)) +
  geom_point(aes(colour = zero), size = 2.5) +
  scale_colour_manual(values = c("FALSE" = "#2166ac", "TRUE" = "#d6604d"),
                      labels = c("non-zero frequency", "zero (constant term)"),
                      name = NULL) +
  scale_x_continuous(breaks = -K:K) +
  scale_y_continuous(breaks = -K:K) +
  coord_equal() +
  labs(title    = sprintf("Frequency grid: K = %d, %d pairs total", K, nrow(freqs)),
       subtitle = "Each point is one (kx, ky) pair — number of cycles across the domain",
       x = "kx  (cycles across domain in x)",
       y = "ky  (cycles across domain in y)") +
  theme_vignette()
```

The red dot at the origin ($k_x = k_y = 0$) is the constant term. The point
at $(1, 0)$ is one full wave across the domain in $x$; $(2, 0)$ is two full
waves; and so on. Negative $k$ values give the sine terms (they are needed
because $\cos(-\omega s) = \cos(\omega s)$ but $\sin(-\omega s) =
-\sin(\omega s)$, so both signs are required to span the full basis).

---

## Step 3: compute A

```{r compute-A}
A <- fourier_integrate_basis(soundings_sf, freqs)

cat("Dimensions of A:", nrow(A), "x", ncol(A), "\n")
cat("  =", n_soundings, "soundings x", 2 * nrow(freqs), "basis columns\n")
cat("  (", nrow(freqs), "frequencies x 2 for cos and sin)\n")
```

Each row of $A$ is one sounding. Each column pair is one frequency — the
odd-indexed column is the cosine integral, the even-indexed is the sine
integral.

Let's look at the row corresponding to the zero frequency, which should
always be 1 (cosine) and 0 (sine):

```{r check-zero-freq}
zero_row <- which(freqs[, 1] == 0 & freqs[, 2] == 0)
cos_col  <- 2 * zero_row - 1
sin_col  <- 2 * zero_row

cat("Zero-frequency column indices: cos =", cos_col, ", sin =", sin_col, "\n\n")
cat("A[, cos_col] (should be 1 for all soundings):\n")
print(round(A[, cos_col], 10))
cat("\nA[, sin_col] (should be 0 for all soundings):\n")
print(round(A[, sin_col], 10))
```

Good — the zero-frequency cos column is exactly 1 everywhere, and the sin
column is exactly 0. This is as expected: the area-average of a constant
function over any polygon is 1.

---

## Step 4: visualise what A looks like

Each row of $A$ encodes how much of each frequency is "seen" by that sounding.
Let's look at how a few specific frequencies vary across soundings.

```{r plot-A-rows, fig.height = 5}
# Pick a low-frequency and a higher-frequency row
# Lowest nonzero x-frequency: omega_x = 2*pi/200km
low_freq_row  <- which(freqs[, 1] == min(freqs[freqs[, 1] > 0, 1]) &
                         freqs[, 2] == 0)
# A mid-frequency
mid_freq_row  <- which(freqs[, 1] == 0 &
                         freqs[, 2] == sort(unique(freqs[freqs[,2]>0, 2]))[3])

sounding_centroids <- sf::st_coordinates(sf::st_centroid(soundings_sf))

plot_df <- data.frame(
  sounding  = seq_len(n_soundings),
  northing  = sounding_centroids[, 2] / 1e3,
  low_cos   = A[, 2 * low_freq_row - 1],
  low_sin   = A[, 2 * low_freq_row],
  mid_cos   = A[, 2 * mid_freq_row - 1],
  mid_sin   = A[, 2 * mid_freq_row]
)

p1 <- ggplot(plot_df, aes(y = northing)) +
  geom_line(aes(x = low_cos, colour = "cos"), linewidth = 0.8) +
  geom_line(aes(x = low_sin, colour = "sin"), linewidth = 0.8) +
  geom_point(aes(x = low_cos, colour = "cos"), size = 2) +
  geom_point(aes(x = low_sin, colour = "sin"), size = 2) +
  scale_colour_manual(values = c(cos = "#2166ac", sin = "#d6604d"), name = NULL) +
  labs(title = "Low y-frequency", x = "A value", y = "Sounding northing (km)") +
  theme_vignette()

p2 <- ggplot(plot_df, aes(y = northing)) +
  geom_line(aes(x = mid_cos, colour = "cos"), linewidth = 0.8) +
  geom_line(aes(x = mid_sin, colour = "sin"), linewidth = 0.8) +
  geom_point(aes(x = mid_cos, colour = "cos"), size = 2) +
  geom_point(aes(x = mid_sin, colour = "sin"), size = 2) +
  scale_colour_manual(values = c(cos = "#2166ac", sin = "#d6604d"), name = NULL) +
  labs(title = "Mid y-frequency", x = "A value", y = "Sounding northing (km)") +
  theme_vignette()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The low-frequency column varies slowly along the orbit track (smoothly across
soundings). The higher-frequency column oscillates faster. This is exactly what
you'd expect: the $A$ matrix encodes how much of each spatial frequency each
sounding "sees", and high-frequency basis functions oscillate rapidly in space.

---

## Step 5: verify by brute-force numerical integration

The integrals are exact analytically, but it's worth checking against a dense
numerical grid for a few soundings and a few frequencies.

```{r verify}
# For each sounding, compare A values against a dense grid mean
verify_one <- function(i, freq_row, N = 500) {
  # Bounding box of sounding i
  bb  <- sf::st_bbox(soundings_sf[i, ])
  xs  <- seq(bb["xmin"], bb["xmax"], length.out = N)
  ys  <- seq(bb["ymin"], bb["ymax"], length.out = N)
  grd <- expand.grid(x = xs, y = ys)

  # Keep only points inside the footprint
  pts_sfc <- sf::st_as_sf(grd, coords = c("x", "y"), crs = 32619)
  inside  <- as.vector(sf::st_within(pts_sfc, soundings_sf[i, ], sparse = FALSE))
  if (sum(inside) == 0) return(c(cos_brute = NA, sin_brute = NA))

  pts_in <- as.matrix(grd[inside, ])
  wx <- freqs[freq_row, 1]
  wy <- freqs[freq_row, 2]
  phase <- pts_in[, 1] * wx + pts_in[, 2] * wy
  c(cos_brute = mean(cos(phase)),
    sin_brute = mean(sin(phase)))
}

# Check a low and a mid frequency for all soundings
check_freq <- low_freq_row

brute <- t(sapply(seq_len(n_soundings), verify_one, freq_row = check_freq))

verify_df <- data.frame(
  sounding  = seq_len(n_soundings),
  cos_exact = round(A[, 2 * check_freq - 1], 6),
  cos_brute = round(brute[, "cos_brute"],    6),
  sin_exact = round(A[, 2 * check_freq],     6),
  sin_brute = round(brute[, "sin_brute"],    6)
)
verify_df$cos_error <- abs(verify_df$cos_exact - verify_df$cos_brute)
verify_df$sin_error <- abs(verify_df$sin_exact - verify_df$sin_brute)

knitr::kable(verify_df,
             caption = paste("Exact vs brute-force grid (N=500) for frequency",
                             check_freq),
             digits = 6)
```

The errors are at the level of the brute-force grid's own discretization error
($\sim 1/N$), not from the exact integration — the analytic formula is exact up
to floating-point precision.

---

## Step 6: SAR-equivalent parameters

The Fourier basis is most useful when it approximates a SAR or Matérn field.
`fourier_prior.R` provides `matern_to_sar()` and `sar_to_matern()` to convert
between the two parameterizations, using the exact correspondence between the
discrete Laplacian and the SPDE $(\kappa^2 - \Delta)u = \mathcal{W}$.

Suppose we want a prior that matches a SAR model on a 5km grid with $\rho =
0.95$:

```{r sar-to-matern}
delta <- 5000   # 5km grid spacing in meters
rho   <- 0.95
tau2  <- 1e6    # some scale

matern_equiv <- sar_to_matern(rho = rho, tau2 = tau2, delta = delta)

cat("SAR parameters:\n")
cat("  rho   =", rho, "\n")
cat("  tau2  =", tau2, "\n")
cat("  delta =", delta, "m\n\n")
cat("Equivalent Matern(nu=1) parameters:\n")
cat("  range  =", round(matern_equiv$range / 1e3, 2), "km\n")
cat("  sigma2 =", round(matern_equiv$sigma2, 4), "\n")
```

And in the other direction — given a fitted Matérn range, what SAR parameters
does it correspond to on a 5km grid?

```{r matern-to-sar}
range  <- 80e3   # 80km range
sigma2 <- 2.0

sar_equiv <- matern_to_sar(range = range, sigma2 = sigma2, delta = delta)

cat("Matern(nu=1) parameters:\n")
cat("  range  =", range / 1e3, "km\n")
cat("  sigma2 =", sigma2, "\n\n")
cat("Equivalent SAR parameters (delta =", delta / 1e3, "km):\n")
cat("  rho  =", round(sar_equiv$rho,  4), "\n")
cat("  tau2 =", round(sar_equiv$tau2, 4), "\n")
```

Let's verify the round-trip:

```{r round-trip}
back <- sar_to_matern(sar_equiv$rho, sar_equiv$tau2, delta)

cat("Round-trip check:\n")
cat("  range  : ", range / 1e3,  "km  ->  rho =", round(sar_equiv$rho, 4),
    "  ->  range:", round(back$range / 1e3, 4), "km\n")
cat("  sigma2 :", sigma2, "  ->  tau2 =", round(sar_equiv$tau2, 2),
    "  ->  sigma2:", round(back$sigma2, 6), "\n")
```

Exact round-trip, as expected from the algebraic derivation.

---

## Summary

Here is the complete workflow in one block:

```{r summary-block, eval = FALSE}
library(spatintegrate)
library(sf)

# 1. Your soundings as an sf object in a projected CRS
soundings_sf <- ...   # sf object, one row per footprint

# 2. Choose a domain and frequency resolution
bbox  <- c(xmin, ymin, xmax, ymax)   # in CRS units (meters)
freqs <- fourier_freq_grid(K = 10, domain_bbox = bbox)

# 3. Compute the A matrix — exact, no quadrature error
A <- fourier_integrate_basis(soundings_sf, freqs)
# A is n_soundings x 2*nrow(freqs)
# column pairs: (cos_f1, sin_f1, cos_f2, sin_f2, ...)

# 4. Interpret SAR <-> Matern parameters
matern_to_sar(range = 80e3, sigma2 = 2, delta = 5000)
sar_to_matern(rho   = 0.95, tau2   = 1e6, delta = 5000)
```

The $A$ matrix is the input to `fastblm` for model fitting. The prior $Q$ is
constructed separately via `fourier_matern_Q_fun()` or `fourier_sar_Q_fun()`,
which are covered in the modeling vignette.
