---
title: "Layer 1: QMC Sampling Primitives"
description: >
  Visual walkthrough of the two Layer 1 functions in spatintegrate:
  generate_qmc_unit_square() and map_unit_square_to_triangle().
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layer 1: QMC Sampling Primitives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 150
)
library(spatintegrate)
library(sf)
library(ggplot2)
library(dplyr)
library(gridExtra)

theme_article <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.border    = element_rect(fill = NA, colour = "grey60"),
      legend.position = "none"
    )
}

region_colours  <- c("#2166ac", "#d6604d", "#4dac26", "#8073ac")
stratum_colours <- c(A = "#a6cee3", B = "#b2df8a")

# Helper: triangulate an sfc polygon into an sfc of triangles.
# Uses centroid filtering to discard Delaunay triangles that fall outside
# the original polygon — necessary for non-convex shapes. This is a
# vignette convenience; the real spatintegrate::triangulate_sf() will
# bake this guarantee into its contract.
triangulate_polygon <- function(poly_sfc) {
  tris <- st_triangulate(poly_sfc) |>
    st_collection_extract("POLYGON") |>
    st_cast("POLYGON")

  # Delaunay triangulation is unconstrained — for non-convex polygons some
  # triangles extend outside the boundary. Filter to interior triangles only.
  centroids <- st_centroid(tris)
  inside    <- st_within(centroids, poly_sfc, sparse = FALSE)[, 1]
  tris[inside]
}

# Helper: extract 3x2 coordinate matrix from a triangle sfg
get_triangle_coords <- function(tri_sfg) {
  coords <- st_coordinates(tri_sfg)
  ext    <- coords[coords[, "L2"] == 1, c("X", "Y"), drop = FALSE]
  if (nrow(ext) >= 2 && all(ext[1, ] == ext[nrow(ext), ])) {
    ext <- ext[-nrow(ext), , drop = FALSE]
  }
  unname(ext)
}

# Helper: sample n_per_triangle QMC points in each triangle of an sfc,
# returning a data frame with x, y, and triangle index.
# This is the per-triangle loop that integrate_one_sounding() will use.
sample_triangulation <- function(tris, n_per_triangle, min_area = 0, id = NULL) {
  areas <- as.numeric(st_area(tris))
  keep  <- areas > min_area
  qmc   <- generate_qmc_unit_square(n_per_triangle)

  do.call(bind_rows, lapply(seq_along(tris)[keep], function(i) {
    pts <- map_unit_square_to_triangle(qmc, get_triangle_coords(tris[[i]]))
    data.frame(x = pts[, 1], y = pts[, 2],
               tri = i, id = if (is.null(id)) NA else id)
  }))
}
```

Layer 1 of `spatintegrate` has exactly two functions:

- `generate_qmc_unit_square(n)` — generates `n` Sobol points in $[0,1]^2$
- `map_unit_square_to_triangle(qmc_points, triangle_coords)` — maps those points into a triangle

Everything else — triangulation, area weighting, basis evaluation — happens in
higher layers. These two functions are pure math with no sf dependency.

---

## 1. `generate_qmc_unit_square()`

A Sobol sequence in $[0,1]^2$. The key property is **low discrepancy** —
points fill space far more evenly than random sampling, which means faster
integration convergence for the same number of points.

```{r qmc-unit-square, fig.height = 3.5}
make_pts_df <- function(n, label) {
  pts <- generate_qmc_unit_square(n)
  data.frame(x = pts[, 1], y = pts[, 2], label = label)
}

set.seed(42)
pts_all <- bind_rows(
  make_pts_df(10,  "Sobol  n = 10"),
  make_pts_df(64,  "Sobol  n = 64"),
  make_pts_df(256, "Sobol  n = 256"),
  data.frame(x = runif(256), y = runif(256), label = "Random  n = 256")
)

sq_outline <- data.frame(x = c(0, 1, 1, 0, 0), y = c(0, 0, 1, 1, 0))

ggplot(pts_all, aes(x, y)) +
  geom_path(data = sq_outline, aes(x, y), colour = "grey40", linewidth = 0.6) +
  geom_point(size = 1.2, colour = "#2166ac") +
  facet_wrap(~label, nrow = 1) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "generate_qmc_unit_square()",
    subtitle = "Low-discrepancy Sobol points vs uniform random",
    x = NULL, y = NULL
  ) +
  theme_article()
```

Even at `n = 10` the Sobol points cover the space without clustering. Random
sampling at `n = 256` still shows visible gaps and clumps.

---

## 2. `map_unit_square_to_triangle()`

Points from $[0,1]^2$ are mapped into a triangle using a barycentric
transformation. Points outside the standard simplex ($u + v > 1$) are reflected
across the diagonal — making the mapping area-preserving and bijective.

```{r map-triangle, fig.height = 4}
qmc64 <- generate_qmc_unit_square(64)

tri_coords  <- matrix(c(0, 0,  2, 0,  0.6, 1.4), ncol = 2, byrow = TRUE)
tri_outline <- data.frame(
  x = c(tri_coords[, 1], tri_coords[1, 1]),
  y = c(tri_coords[, 2], tri_coords[1, 2])
)

mapped      <- map_unit_square_to_triangle(qmc64, tri_coords)
tri_samples <- data.frame(x = mapped[, 1], y = mapped[, 2])

p_tri_geom <- ggplot() +
  geom_polygon(data = tri_outline, aes(x, y),
               fill = "#2166ac", alpha = 0.15, colour = "black", linewidth = 0.7) +
  labs(title = "Triangle polygon", x = NULL, y = NULL) +
  theme_article()

p_tri_samp <- ggplot() +
  geom_polygon(data = tri_outline, aes(x, y),
               fill = NA, colour = "black", linewidth = 0.5, linetype = "dashed") +
  geom_point(data = tri_samples, aes(x, y),
             size = 1.5, colour = "#2166ac", alpha = 0.8) +
  labs(title = "QMC samples (64 points)", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_tri_geom, p_tri_samp, ncol = 2)
```

Points fill the triangle uniformly. The same `qmc64` sequence can be reused
across multiple triangles — each independent barycentric mapping produces a
uniform sample inside its own triangle.

---

## How these two functions compose

In practice, `integrate_one_sounding()` (Layer 4) calls them like this:

```r
qmc   <- generate_qmc_unit_square(n_per_triangle)   # once per sounding

for each triangle i in triangulation:
  pts_i  <- map_unit_square_to_triangle(qmc, tri_coords_i)
  phi_i  <- basis_fn(pts_i)                          # evaluated, then discarded
  A_row  <- A_row + weight_i * colMeans(phi_i)
```

One QMC sequence, reused. One basis evaluation per triangle, immediately
discarded. Memory stays bounded at `n_per_triangle × k` regardless of how
many triangles the sounding has.

The four use cases below show how the geometry preparation feeds into this loop.

---

## Use Case 1: Overlapping Quadrilaterals

Satellite footprints are near-parallelogram quadrilaterals. We triangulate each
one into two triangles and call `map_unit_square_to_triangle` on each.

```{r uc1-quads, fig.height = 4}
make_parallelogram <- function(origin, width, height, shear) {
  x0 <- origin[1]; y0 <- origin[2]
  matrix(c(
    x0,                  y0,
    x0 + width,          y0,
    x0 + width + shear,  y0 + height,
    x0 + shear,          y0 + height
  ), ncol = 2, byrow = TRUE)
}

footprints <- list(
  make_parallelogram(c(0.0, 0.0), 2, 1, 0.3),
  make_parallelogram(c(0.5, 0.2), 2, 1, 0.3),
  make_parallelogram(c(1.0, 0.4), 2, 1, 0.3),
  make_parallelogram(c(0.3, 0.7), 2, 1, 0.3)
)

fp_sfc_list <- lapply(footprints, function(f) {
  st_sfc(st_polygon(list(rbind(f, f[1, ]))))
})

fp_tri_list <- lapply(fp_sfc_list, triangulate_polygon)

fp_outlines <- do.call(bind_rows, lapply(seq_along(footprints), function(i) {
  coords <- rbind(footprints[[i]], footprints[[i]][1, ])
  data.frame(x = coords[, 1], y = coords[, 2], id = factor(i))
}))

fp_samples <- do.call(bind_rows, lapply(seq_along(fp_tri_list), function(i) {
  sample_triangulation(fp_tri_list[[i]], n_per_triangle = 32) |>
    mutate(id = factor(i))
}))

p_geom <- ggplot() +
  geom_polygon(data = fp_outlines, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.15, linewidth = 0.7) +
  scale_fill_manual(values   = region_colours) +
  scale_colour_manual(values = region_colours) +
  labs(title = "Footprint polygons", x = NULL, y = NULL) +
  theme_article()

p_samp <- ggplot() +
  geom_polygon(data = fp_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.5, linetype = "dashed") +
  geom_point(data = fp_samples, aes(x, y, colour = id), size = 1.2, alpha = 0.8) +
  scale_colour_manual(values = region_colours) +
  labs(title = "QMC samples (32 per triangle)", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_geom, p_samp, ncol = 2)
```

Each parallelogram becomes two triangles. Each triangle gets 32 independent
QMC points. The overlapping regions are resolved at the integration layer above.

---

## Use Case 2: Irregular Polygons

Irregular polygons triangulate into more pieces, producing more triangles and
therefore more sample points — denser sampling emerges naturally from finer
triangulation.

```{r uc2-irregular, fig.height = 4.5}
make_blob <- function(cx, cy, n_pts = 7, seed = 1) {
  set.seed(seed)
  angles <- sort(runif(n_pts, 0, 2 * pi))
  r      <- 0.5 + runif(n_pts, 0, 0.4)
  xs     <- cx + r * cos(angles)
  ys     <- cy + r * sin(angles)
  st_polygon(list(cbind(c(xs, xs[1]), c(ys, ys[1]))))
}

blobs <- st_sfc(
  make_blob(0,   0,   seed = 1),
  make_blob(2,   0.3, seed = 5),
  make_blob(1,   1.5, seed = 9)
)

tri_list <- lapply(blobs, function(b) triangulate_polygon(st_sfc(b)))

blob_outlines <- do.call(bind_rows, lapply(seq_along(blobs), function(i) {
  coords <- st_coordinates(blobs[i])
  data.frame(x = coords[, "X"], y = coords[, "Y"], id = factor(i))
}))

tri_outlines <- do.call(bind_rows, lapply(seq_along(tri_list), function(i) {
  do.call(bind_rows, lapply(seq_along(tri_list[[i]]), function(j) {
    coords <- st_coordinates(tri_list[[i]][j])
    data.frame(x = coords[, "X"], y = coords[, "Y"],
               blob = factor(i), tri = j)
  }))
}))

blob_samples <- do.call(bind_rows, lapply(seq_along(tri_list), function(i) {
  sample_triangulation(tri_list[[i]], n_per_triangle = 16) |>
    mutate(id = factor(i))
}))

p_blobs <- ggplot() +
  geom_polygon(data = blob_outlines, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.15, linewidth = 0.8) +
  scale_fill_manual(values = region_colours) +
  scale_colour_manual(values = region_colours) +
  labs(title = "Irregular polygons", x = NULL, y = NULL) +
  theme_article()

p_tris <- ggplot() +
  geom_polygon(data = blob_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.9) +
  geom_polygon(data = tri_outlines,
               aes(x, y, group = interaction(blob, tri)),
               fill = NA, colour = "grey50", linewidth = 0.3) +
  scale_colour_manual(values = region_colours) +
  labs(title = "Triangulated", x = NULL, y = NULL) +
  theme_article()

p_samp2 <- ggplot() +
  geom_polygon(data = blob_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.5, linetype = "dashed") +
  geom_point(data = blob_samples, aes(x, y, colour = id),
             size = 1.0, alpha = 0.8) +
  scale_colour_manual(values = region_colours) +
  labs(title = "QMC samples (16 per triangle)", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_blobs, p_tris, p_samp2, ncol = 3)
```

The middle panel shows the triangulation explicitly. More triangles = more
points = higher effective sampling density, without any special budget logic.

---

## Use Case 3: Quadrilaterals with Strata

When basis functions have discontinuities at known boundaries, we intersect
each footprint with the stratum polygons before triangulating. This guarantees
no QMC point ever straddles a boundary.

```{r uc3-quad-strata, fig.height = 4.5, fig.width = 9}
stratum_A <- st_sfc(st_polygon(list(matrix(
  c(1.0, 0.3, 1.6, 0.1, 2.3, 0.4, 2.5, 0.9,
    2.1, 1.3, 1.4, 1.5, 0.9, 1.1, 0.8, 0.6, 1.0, 0.3),
  ncol = 2, byrow = TRUE
))))

fp_sfc <- st_sfc(lapply(footprints, function(f) {
  st_polygon(list(rbind(f, f[1, ])))
}))

domain    <- st_as_sfc(st_bbox(c(xmin = -0.5, ymin = -0.5, xmax = 4, ymax = 2.5)))
stratum_B <- st_difference(domain, stratum_A)
strata    <- list(A = stratum_A, B = stratum_B)

intersect_with_strata <- function(poly, strata) {
  lapply(strata, function(s) {
    inter <- suppressWarnings(st_intersection(poly, s))
    if (length(inter) == 0 || st_is_empty(inter)) return(NULL)
    inter
  })
}

fp_pieces <- lapply(seq_along(fp_sfc), function(i) {
  intersect_with_strata(fp_sfc[i], strata)
})

fp_piece_tris <- lapply(fp_pieces, function(pieces) {
  lapply(pieces, function(piece) {
    if (is.null(piece)) return(NULL)
    triangulate_polygon(piece)
  })
})

stratum_A_coords <- st_coordinates(stratum_A)
stratum_A_df     <- data.frame(x = stratum_A_coords[, "X"],
                                y = stratum_A_coords[, "Y"])

piece_outlines <- do.call(bind_rows, lapply(seq_along(fp_pieces), function(i) {
  do.call(bind_rows, lapply(names(strata), function(sname) {
    piece <- fp_pieces[[i]][[sname]]
    if (is.null(piece)) return(NULL)
    coords <- st_coordinates(piece)
    data.frame(x = coords[, "X"], y = coords[, "Y"],
               fp = factor(i), stratum = sname,
               group = paste(i, sname))
  }))
}))

piece_samples <- do.call(bind_rows, lapply(seq_along(fp_piece_tris), function(i) {
  do.call(bind_rows, lapply(names(strata), function(sname) {
    tris <- fp_piece_tris[[i]][[sname]]
    if (is.null(tris) || length(tris) == 0) return(NULL)
    sample_triangulation(tris, n_per_triangle = 16) |>
      mutate(fp = factor(i), stratum = sname)
  }))
}))

xlims <- c(-0.2, 3.8); ylims <- c(-0.2, 2.0)

p_fp_only <- ggplot() +
  geom_polygon(data = fp_outlines, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.15, linewidth = 0.7) +
  scale_fill_manual(values = region_colours) +
  scale_colour_manual(values = region_colours) +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  labs(title = "Footprints", x = NULL, y = NULL) +
  theme_article()

p_strata_only <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.7,
               colour = "grey30", linewidth = 0.7) +
  annotate("text", x = 1.65, y = 0.8, label = "A",
           size = 5, fontface = "bold", colour = "grey20") +
  annotate("text", x = 3.2,  y = 1.7, label = "B",
           size = 5, fontface = "bold", colour = "grey50") +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  labs(title = "Strata", x = NULL, y = NULL) +
  theme_article()

p_combined <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.5, colour = NA) +
  geom_polygon(data = fp_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.7) +
  scale_colour_manual(values = region_colours) +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  labs(title = "Combined", x = NULL, y = NULL) +
  theme_article()

p_samples <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.3, colour = NA) +
  geom_polygon(data = fp_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.4, linetype = "dashed") +
  geom_point(data = piece_samples, aes(x, y, colour = fp, shape = stratum),
             size = 1.3, alpha = 0.85) +
  scale_colour_manual(values = region_colours) +
  scale_shape_manual(values  = c(A = 17, B = 16)) +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  labs(title = "QMC samples (▲ = A, ● = B)", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_fp_only, p_strata_only, p_combined, p_samples, ncol = 4)
```

Stratum A is small — fewer triangles fall in it, so it gets fewer points
naturally. No point crosses the A/B boundary because the intersection step
happens before triangulation.

---

## Use Case 4: Irregular Polygons with Strata

The same pipeline as UC3, but the observation polygons are irregular blobs
rather than parallelograms. The code is identical — shape complexity is
absorbed by triangulation.

```{r uc4-irregular-strata, fig.height = 4.5, fig.width = 9}
blob_pieces <- lapply(seq_along(blobs), function(i) {
  intersect_with_strata(blobs[i], strata)
})

blob_piece_tris <- lapply(blob_pieces, function(pieces) {
  lapply(pieces, function(piece) {
    if (is.null(piece)) return(NULL)
    triangulate_polygon(piece)
  })
})

blob_piece_outlines <- do.call(bind_rows, lapply(seq_along(blob_pieces), function(i) {
  do.call(bind_rows, lapply(names(strata), function(sname) {
    piece <- blob_pieces[[i]][[sname]]
    if (is.null(piece)) return(NULL)
    coords <- st_coordinates(piece)
    data.frame(x = coords[, "X"], y = coords[, "Y"],
               blob = factor(i), stratum = sname,
               group = paste(i, sname))
  }))
}))

blob_piece_samples <- do.call(bind_rows, lapply(seq_along(blob_piece_tris), function(i) {
  do.call(bind_rows, lapply(names(strata), function(sname) {
    tris <- blob_piece_tris[[i]][[sname]]
    if (is.null(tris) || length(tris) == 0) return(NULL)
    sample_triangulation(tris, n_per_triangle = 16) |>
      mutate(blob = factor(i), stratum = sname)
  }))
}))

p_fp4_only <- ggplot() +
  geom_polygon(data = blob_outlines, aes(x, y, fill = id, colour = id, group = id),
               alpha = 0.15, linewidth = 0.8) +
  scale_fill_manual(values = region_colours) +
  scale_colour_manual(values = region_colours) +
  labs(title = "Irregular polygons", x = NULL, y = NULL) +
  theme_article()

p_strata4_only <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.7,
               colour = "grey30", linewidth = 0.7) +
  annotate("text", x = 1.65, y = 0.8, label = "A",
           size = 5, fontface = "bold", colour = "grey20") +
  annotate("text", x = 3.2,  y = 1.7, label = "B",
           size = 5, fontface = "bold", colour = "grey50") +
  labs(title = "Strata", x = NULL, y = NULL) +
  theme_article()

p_combined4 <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.5, colour = NA) +
  geom_polygon(data = blob_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.7) +
  scale_colour_manual(values = region_colours) +
  labs(title = "Combined", x = NULL, y = NULL) +
  theme_article()

p_samp4 <- ggplot() +
  geom_polygon(data = stratum_A_df, aes(x, y),
               fill = stratum_colours["A"], alpha = 0.3, colour = NA) +
  geom_polygon(data = blob_outlines, aes(x, y, colour = id, group = id),
               fill = NA, linewidth = 0.4, linetype = "dashed") +
  geom_point(data = blob_piece_samples,
             aes(x, y, colour = blob, shape = stratum),
             size = 1.3, alpha = 0.85) +
  scale_colour_manual(values = region_colours) +
  scale_shape_manual(values  = c(A = 17, B = 16)) +
  labs(title = "QMC samples (▲ = A, ● = B)", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_fp4_only, p_strata4_only, p_combined4, p_samp4, ncol = 4)
```

Identical pipeline to UC3. The irregular shapes produce more triangles in some
regions, giving them higher sampling density automatically.

---

## Summary

Layer 1 is just two functions. Everything else is geometry preparation.

| Step | Function | Layer |
|---|---|---|
| Generate QMC points | `generate_qmc_unit_square(n)` | 1 |
| Map points into triangle | `map_unit_square_to_triangle(pts, coords)` | 1 |
| Triangulate a polygon | `triangulate_sf()` | 2 |
| Intersect with strata | `st_intersection()` | 2 |
| Loop, weight, accumulate | `integrate_one_sounding()` | 4 |

The four use cases above show how UC3 and UC4 differ from UC1 and UC2 only in
the geometry preparation step — the Layer 1 calls are identical in all cases.
