---
title: "Layer 2: Polygon Triangulation"
description: >
  How spatintegrate decomposes polygons into triangles for QMC integration —
  covering triangulate_sf(), get_triangle_coords(), and the CRS requirement.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layer 2: Polygon Triangulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dpi        = 150
)
library(spatintegrate)
library(sf)
library(ggplot2)
library(dplyr)
library(gridExtra)

# Projected CRS used throughout (UTM zone 19N, Boston area)
TEST_CRS <- 32619

theme_article <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.border    = element_rect(fill = NA, colour = "grey60"),
      legend.position = "none"
    )
}

polygon_to_df <- function(sfc) {
  coords <- sf::st_coordinates(sfc)
  data.frame(x = coords[, "X"], y = coords[, "Y"])
}

triangles_to_df <- function(tris) {
  do.call(bind_rows, lapply(seq_along(tris), function(i) {
    coords <- sf::st_coordinates(tris[i])
    data.frame(x = coords[, "X"], y = coords[, "Y"], tri = factor(i))
  }))
}
```

Layer 2 of `spatintegrate` handles geometry preparation: turning polygons into
triangles that Layer 1 can map QMC points into. Two functions:

- `triangulate_sf(x)` — decompose a polygon into interior triangles
- `get_triangle_coords(x)` — extract a 3×2 coordinate matrix from a triangle

---

## The CRS requirement

**All inputs must be in a projected CRS.** This is a hard requirement, not a
convenience. The barycentric mapping in `map_unit_square_to_triangle()` is
purely Euclidean — it treats coordinates as flat (x, y) pairs. In a geographic
CRS (lon/lat), a degree of longitude is not the same physical distance as a
degree of latitude, and both vary with position. The mapping would produce
points that are uniform in degree-space but not in physical space, silently
biasing your integrals.

```{r crs-error}
poly_geo <- sf::st_sfc(
  sf::st_polygon(list(matrix(c(0,0,1,0,1,1,0,1,0,0), ncol = 2, byrow = TRUE))),
  crs = 4326
)
tryCatch(
  triangulate_sf(poly_geo),
  error = function(e) cat("Error:", conditionMessage(e), "\n")
)
```

Use `ensure_projected()` to convert before calling:

```{r ensure-projected, eval = FALSE}
poly_proj <- ensure_projected(poly_geo)
tris      <- triangulate_sf(poly_proj)
```

All examples below use UTM zone 19N (EPSG:32619) — projected, in meters.

---

## 1. `triangulate_sf()`

### How it works

`triangulate_sf()` uses **constrained Delaunay triangulation** when GEOS ≥ 3.10
is available (the default on current installations). Constrained triangulation
respects polygon boundary edges — triangles cannot cross the boundary, so all
output triangles are interior by construction with no post-processing needed.

```{r geos-version}
geos_ver <- sf::sf_extSoftVersion()[["GEOS"]]
cat("GEOS version:", geos_ver, "\n")
cat("Using constrained triangulation:",
    utils::compareVersion(geos_ver, "3.10.0") >= 0, "\n")
```

On older GEOS installations the function falls back to unconstrained Delaunay
plus a centroid filter — see the quality note at the end.

### Convex polygons

A convex polygon with $n$ vertices produces $n - 2$ triangles.

```{r convex, fig.height = 4}
para <- sf::st_sfc(sf::st_polygon(list(matrix(
  c(0,0, 3,0, 3.5,2, 0.5,2, 0,0), ncol = 2, byrow = TRUE
))), crs = TEST_CRS)

tris <- triangulate_sf(para)
cat("Triangles:", length(tris), "\n")

p1 <- ggplot() +
  geom_polygon(data = polygon_to_df(para), aes(x, y),
               fill = "#2166ac", alpha = 0.15,
               colour = "black", linewidth = 0.8) +
  coord_equal() +
  labs(title = "Parallelogram", x = NULL, y = NULL) +
  theme_article()

p2 <- ggplot() +
  geom_polygon(data = triangles_to_df(tris),
               aes(x, y, fill = tri, group = tri),
               alpha = 0.3, colour = "black", linewidth = 0.6) +
  scale_fill_manual(values = c("#2166ac", "#d6604d")) +
  coord_equal() +
  labs(title = "2 interior triangles", x = NULL, y = NULL) +
  theme_article()

grid.arrange(p1, p2, ncol = 2)
```

### Non-convex polygons

This is where the constrained vs unconstrained distinction matters. Unconstrained
Delaunay triangulation fills concave regions with exterior triangles —
`triangulate_sf()` avoids this by using constrained triangulation.

```{r nonconvex, fig.height = 4}
l_shape <- sf::st_sfc(sf::st_polygon(list(matrix(
  c(0,0, 3,0, 3,1, 1,1, 1,3, 0,3, 0,0), ncol = 2, byrow = TRUE
))), crs = TEST_CRS)

# Unconstrained: fills the concave notch with exterior triangles
tris_unconstrained <- sf::st_triangulate(l_shape) |>
  sf::st_collection_extract("POLYGON") |>
  sf::st_cast("POLYGON")

# triangulate_sf: interior triangles only
tris_constrained <- triangulate_sf(l_shape)

cat("Unconstrained triangles:", length(tris_unconstrained), "\n")
cat("triangulate_sf result: ", length(tris_constrained),   "\n")

poly_df <- polygon_to_df(l_shape)

p_unc <- ggplot() +
  geom_polygon(data = poly_df, aes(x, y),
               fill = NA, colour = "black", linewidth = 0.9) +
  geom_polygon(data = triangles_to_df(tris_unconstrained),
               aes(x, y, fill = tri, group = tri),
               alpha = 0.25, colour = "grey40", linewidth = 0.4) +
  coord_equal() +
  labs(title    = "Unconstrained Delaunay",
       subtitle = paste(length(tris_unconstrained), "triangles — includes exterior"),
       x = NULL, y = NULL) +
  theme_article()

p_con <- ggplot() +
  geom_polygon(data = poly_df, aes(x, y),
               fill = NA, colour = "black", linewidth = 0.9) +
  geom_polygon(data = triangles_to_df(tris_constrained),
               aes(x, y, fill = tri, group = tri),
               alpha = 0.25, colour = "grey40", linewidth = 0.4) +
  coord_equal() +
  labs(title    = "triangulate_sf()",
       subtitle = paste(length(tris_constrained), "triangles — interior only"),
       x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_unc, p_con, ncol = 2)
```

The constrained result covers exactly the polygon interior — verified by area:

```{r area-check}
area_poly <- as.numeric(sf::st_area(l_shape))
area_tris <- sum(as.numeric(sf::st_area(tris_constrained)))
cat("Polygon area:  ", round(area_poly, 4), "\n")
cat("Triangles area:", round(area_tris, 4), "\n")
cat("Match:", isTRUE(all.equal(area_poly, area_tris, tolerance = 1e-6)), "\n")
```

### Multipolygons

MULTIPOLYGON inputs are cast to individual polygons first, then each is
triangulated.

```{r multipolygon}
two_sq <- sf::st_sfc(sf::st_multipolygon(list(
  list(matrix(c(0,0,1,0,1,1,0,1,0,0), ncol=2, byrow=TRUE)),
  list(matrix(c(3,0,4,0,4,1,3,1,3,0), ncol=2, byrow=TRUE))
)), crs = TEST_CRS)

tris_mp <- triangulate_sf(two_sq)
cat("Two unit squares →", length(tris_mp), "triangles\n")  # 4
```

### The `min_area` argument

Floating-point slivers from `sf::st_intersection()` occasionally produce
near-zero-area triangles. Drop them with `min_area` (in CRS units²):

```{r min-area, eval = FALSE}
# Drop triangles smaller than 1 m² (for a UTM projection in meters)
tris <- triangulate_sf(intersection_piece, min_area = 1)
```

The default `min_area = 0` drops only exactly-zero-area triangles.

---

## 2. `get_triangle_coords()`

Given a single triangle sfg or length-1 sfc, returns the three vertex
coordinates as a 3×2 matrix — ready to pass to `map_unit_square_to_triangle()`.

```{r get-coords}
tris   <- triangulate_sf(l_shape)
coords <- get_triangle_coords(tris[[1]])

cat("Dim:", dim(coords), "\n")
print(coords)
```

sf repeats the first vertex to close the polygon ring — `get_triangle_coords()`
strips that duplicate so you always get exactly 3 rows.

---

## Putting it together: Layer 1 + Layer 2

The composition: triangulate with Layer 2, map QMC points with Layer 1.
This is the per-triangle loop that `integrate_one_sounding()` (Layer 4)
will use.

```{r end-to-end, fig.height = 4.5}
# A strongly concave C-shape
c_shape <- sf::st_sfc(sf::st_polygon(list(matrix(
  c(0,0, 4,0, 4,1, 1,1, 1,3, 4,3, 4,4, 0,4, 0,0),
  ncol = 2, byrow = TRUE
))), crs = TEST_CRS)

tris <- triangulate_sf(c_shape)
qmc  <- generate_qmc_unit_square(32)   # generated once, reused per triangle

samples <- do.call(bind_rows, lapply(seq_along(tris), function(i) {
  pts <- map_unit_square_to_triangle(qmc, get_triangle_coords(tris[[i]]))
  data.frame(x = pts[,1], y = pts[,2], tri = factor(i))
}))

p_tris <- ggplot() +
  geom_polygon(data = polygon_to_df(c_shape), aes(x, y),
               fill = NA, colour = "black", linewidth = 0.9) +
  geom_polygon(data = triangles_to_df(tris),
               aes(x, y, fill = tri, group = tri),
               alpha = 0.2, colour = "grey50", linewidth = 0.4) +
  coord_equal() +
  labs(title    = "C-shape triangulated",
       subtitle = paste(length(tris), "interior triangles"),
       x = NULL, y = NULL) +
  theme_article()

p_samp <- ggplot() +
  geom_polygon(data = polygon_to_df(c_shape), aes(x, y),
               fill = NA, colour = "black",
               linewidth = 0.5, linetype = "dashed") +
  geom_point(data = samples, aes(x, y, colour = tri),
             size = 1.3, alpha = 0.8) +
  coord_equal() +
  labs(title    = "32 QMC points per triangle",
       subtitle = paste(nrow(samples), "total"),
       x = NULL, y = NULL) +
  theme_article()

grid.arrange(p_tris, p_samp, ncol = 2)
```

All points fall inside the C-shape — none in the concave notch. Each triangle
gets exactly 32 points regardless of its size: density is controlled by
triangulation fineness, not by a sampling budget.

---

## Quality note

Both constrained and unconstrained Delaunay triangulation **maximise the
minimum angle** across all triangles — the provably optimal shape quality from
a fixed vertex set. Thin sliver triangles can still arise when strata
intersections produce narrow pieces. For minimum-angle guarantees via Steiner
point insertion, `sfdct::ct_triangulate()` (which wraps Shewchuk's Triangle
library) is the upgrade path. The swap would be isolated to `triangulate_sf()`
— nothing else in the stack would change.

---

## Summary

| Function | Input | Output | Key guarantee |
|---|---|---|---|
| `triangulate_sf(x)` | Projected polygon | sfc of triangles | Interior triangles only |
| `get_triangle_coords(x)` | Single triangle sfg/sfc | 3×2 matrix | No closing duplicate |

Both functions require a projected CRS. Use `ensure_projected()` to convert
geographic inputs.
